# Basic Visualization

The R Programming language has some of the most powerful data visualization packages available.  These packages are continually expanded upon, with new data visualizations packages being created on a regular basis.  In addition to packages that create your basic statistical visualization (line plots, bar plots, pie plots, etc) there are packages that create geospatial visualizations, 3D visualizations, as well as interactive visualizations.

Lets start by reading in the Korean Conflict data and perform the primary cleaning functions that we performed last lesson.  

```{r, warning = FALSE}
kor <- read.csv('KoreanConflict.csv', as.is=TRUE)
kor$BIRTH_YEAR <- as.numeric(kor$BIRTH_YEAR)
kor <- kor[!is.na(kor$BIRTH_YEAR),]
```

Now that we have the data in memory, we will use some basic visualizations to explore the data.  In this lesson, we will primarily use visualizations from the ggplot2 package [@R-ggplot2].  If you haven't installed this package yet, run the command `install.packages('ggplot2')`.


##Bar Plot

We will start by producing a basic barplot of categorical variables.  First we'll look at the ``BRANCH`` field for the Korean Casualties.

```{r, warning = FALSE, message = FALSE}
library(ggplot2)
ggplot(kor, aes(BRANCH)) + geom_bar()
```

At first this command isn't very intuitive.  The ``ggplot`` command is used with all visualizations.  In this command it says that we will create a visualization of the ``kor`` data set, and particularly look at the ``BRANCH`` variable.  The next command says to take this specific field and create a barplot (we could create other plots with this data as well.)

If we wanted to improve the color scheme, we could add `fill = BRANCH`.  

```{r}
ggplot(kor, aes(BRANCH, fill = BRANCH)) + geom_bar()

```

Now, let's add a title.  Additionally, we will get rid of the legend, since it the labels are already on the axis.

```{r}
ggplot(kor, aes(BRANCH, fill = BRANCH)) + geom_bar() + 
  ggtitle("Casualties by Service")+ theme(legend.position="none")

```

What if we wanted to stacked barplot?  Say we wanted to see how the distribution of ETHNICITY in the service BRANCHES.  First, let's take a look at our categories:

```{r}
table(kor$ETHNICITY_2)
```

These are rather long names to display on a chart.  We will start by creating shorter names.  We can do this in the code below using the ``grep command that we learned in Lesson 2:

```{r}
kor$ETHNICITY_2[grep("HAWAIIAN",kor$ETHNICITY_2)] <- "HAWAIIAN"

```

Now all we have to do to change in our previous code is to change the ``fill = BRANCH`` to ``fill = ETHNICITY_2``:

```{r}
ggplot(kor, aes(BRANCH, fill = ETHNICITY_2)) + geom_bar() + 
  ggtitle("Casualties by Service")
```

##Pie Plot

Statisticians will generally tell you that you should never use a Pie Plot (usually a bar plot is recommended because it is easier for the human eye to distinguish differences in magnitude).  That being said, there are still a few occasional times when a pie plot is necessary. For this plot, we are going to use a function from the BASE graphics package (this comes with R and you don't have to load it). The pie plot is very easy to produce if we wrap the ``pie()`` command around the ``table`` command:

```{r}
pie(table(kor$BRANCH), main = "Korean War Casualties by Service")

```

##Histogram Plot

Now we'll take a look at several ways to create *histograms* in R.  This is where R will quickly outshine Microsoft Excel and other spreadsheet programs.  While Excel can create a barplot just like R, it is extremely time consuming to create a *histogram* in Excel, whereas R can create one in one line code.  

Let's create a histogram of the age or each Korean Casualty.  Notice that we don't have a field that has *age* in it, but we do have the ``BIRTH_YEAR`` and ``FATALITY_YEAR``.  The calculation below will create a new field that is the ``AGE`` of the casualty at death (notice that we remove all records with a ``FATALITY_DATE`` after 1960 in order to remove those who were MIA and declared dead at a later time).

```{r}
# Filter our MIA
kor2 <- dplyr::filter(kor, FATALITY_DATE < 1960) 

# Coerce to numeric
kor2$FATALITY_YEAR <- as.numeric(kor2$FATALITY_YEAR)  
kor2$BIRTH_YEAR <- as.numeric(kor2$BIRTH_YEAR)

# Create AGE field
kor2$AGE <- kor2$FATALITY_YEAR - kor2$BIRTH_YEAR      

```

Now that we've created the ``AGE`` field, we will use two different techniques to create a histogram.  The Base R package has a histogram function that is easy to use and helpful for exploring data.  

```{r}
hist(kor2$AGE, freq=TRUE, main = "Historgram of Age", xlab = "AGE")

```

The ``ggplot2`` package also has the ability to generate a histogram that is generally better for presentations.  

```{r, warning = FALSE, message = FALSE}
ggplot(kor2, aes(AGE)) + geom_histogram() + ggtitle('Korean Casualty Age Distribution')

```



##Time Series Plot

Time series plots (and line plots in general) are helpful in visually identifying trends and anomalies in data. We are going to change our data sets to look at the Olympic data. This data set is the data on all Olympic metals for the Summer Olympics from 1896 to 2012.  This data set is available here:

[https://s3.amazonaws.com/dscoe-data/summer.csv](	https://s3.amazonaws.com/dscoe-data/summer.csv)

You can download the file with the following command:

```{r, eval = FALSE}
download.file("https://s3.amazonaws.com/dscoe-data/summer.csv", destfile = "summer.csv")

```

Now that you've acquired the data, read it into R and take a look at its structure with the following two commands:

```{r}
oly <- read.csv('summer.csv', as.is = TRUE)
str(oly)

```

Let's say that we want to explore the trend of increased numbers of women participating in the Summer Olympics over the past century.  To do this we could start by aggregating the number of Olympic medals by ``Year`` and ``Gender``: 

```{r}
oly_sum <- aggregate(Athlete ~ Year + Gender, data=oly, length)

```

Let's take a look at the first few lines of this new data set to make sure that it aggregated the data as we anticipated:

```{r}
head(oly_sum)

```


Everything looks good.  Now we have a data set that we can use to create a time series line plot.  We create this plot below:

```{r}
ggplot(oly_sum, aes(Year, Athlete, group = Gender, color = Gender)) + geom_line() + ggtitle('Summer Olympic Metals by Gender')
```

From this graph we see that there was significant growth in the participation of women in the Olympics starting in the 1970's (we also see an interesting spike in the number of medals for men in the 1920's that we could explore if desired).  

Now let's select a few of the prominent countries in the Summer Olympics (we'll look at the permanent members of the UN Security Council: United States, Great Britain, France, China, and Russia).  To start this let's first look at the table of countries listed in the data set.  

```{r}
table(oly$Country)
```

Studying this data a bit, we see that the ISO-3 code for Russia changed from *URS* to *RUS* after the fall of the Soviet Union.  For our analysis, we will change all *USR* data to *RUS*.  

```{r}
oly$Country[oly$Country=="URS"] <- "RUS"

```

Now will will aggregate the data by ``Year`` and ``Country`` as well as filter out the counties we are interested in (notice how we create a vector of our five countries and then use the ``%in%`` function to filter out multiple countries):


```{r, warning=FALSE,message=FALSE}
library(dplyr)
oly_country <- aggregate(Medal ~ Country + Year, data = oly, length)

countries <- c('USA','GBR','FRA','CHN','RUS')
oly_country2 <- filter(oly_country, Country %in% countries)

```

Having completed this, we will now plot a time series plot by country for the last century.  

```{r}
ggplot(oly_country2, aes(Year,Medal,group = Country, color = Country)) + geom_line() + ggtitle('Summer Olympic Metals by Country')
```

This plot highlights a bit of the history of the summer Olympics.  Note that Russia began competing in the 1950's and China didn't begin competing until the mid-1980's.  While the US has the longest sustained volume, Russia has the highest number in a single year (1980).  While Great Britain had a surge at the turn of the century, it has decreased to ~50 medals in the 1920's, and stayed near this mark for most the century (as has France).

## Practice Problem

Using the Olympic Data and Google, try to recreate the plot below with a horizontal barplot and bars ordered by volume of medals.

```{r, echo = FALSE}
oly<- within(oly, Sport <- factor(Sport, levels = names(sort(table(Sport)))))

ggplot(oly, aes(Sport, fill =1)) + geom_bar() + coord_flip() +   theme(legend.position="none") + ggtitle('Summer Olympic Medals by Sport')
```


