# Introduction to Dates in R

We often see dates and times in data.  Often each record (or row) of data is connected to at least one date or time.  Similar to Microsoft Excel, R has a special class or format that it uses to work with dates.  

## Dates with Base R

We will start by showing a few of the date commands that are built into the Base R package (later on we will take a look at the *lubridate* package [@R-lubridate], which has some more user friendly functions.)

First we will demonstrate a couple commands that will generate the current date for your system (either your physical computer or your cloud computer).  Below is the system date:

```{r}
Sys.Date()

```

Next we will show the system time down to hours, minutes, and seconds with the Time Zone:

```{r}
Sys.time()

```

Note that that these are not character objects:

```{r}
class(Sys.Date())

```

This is a special class called the *Date* class.  When you read data into R, your fields that have dates are normally converted to the *character* class, not the *date* class.  In order to convert from a *character* class to the *date* class in Base R, use the code below. 

```{r}
# Create a character vector of random dates
myDates <- c("2016-02-07", "2016-04-02","2016-06-28")

#Convert character vector to dates vector
myDates <- as.Date(myDates)

myDates

```

Now we'll check to make sure we've converted it to the proper class of data:

```{r}
class(myDates)

```

Now that this is a date object, we can conduct mathematical operations that we could not conduct with a character vector, like subtracting 5 days from all dates:

```{r}
myDates - 5

```

or checking the difference between dates:

```{r}
Sys.Date() - myDates[1]

```




The date formatting code above will only work as described if my input data are formatted exactly as shown, with four-digit years, two-digit months and days, and hyphens in between.  In order to convert dates in a different format, you will use the format parameter and describe your unique date format as seen below:


```{r}
# Create a character vector of random dates
myDates <- c("02/07/2016", "04/02/2016","06/28/2016")

#Convert character vector to dates vector
myDates <- as.Date(myDates, format = "%m/%d/%Y")

myDates

```

Below is a table of all the most common date components and their abbreviation.  

Conversion Specification | Definition
-------------------------| ------------------------------------
%a  |   Abbreviated weekday 
%A  |   Full weekday 
%b  |   Abbreviated month 
%B  |   Full month 
%d  |   Day of the month as decimal number (01–31).
%H  |   Hours as decimal number (00–23)
%I  |  Hours as decimal number (01–12)
%m  |  Month as decimal number (01–12)
%M  |  Minute as decimal number (00–59)
%p  |  AM/PM indicator in the locale. Used in conjunction with %I and not with %H
%S  |  Second as integer (00–61), allowing for up to two leap-seconds 
%w  |  Weekday as decimal number (0–6, Sunday is 0).
%y  |  Year with two digits (87)
%Y  |  Year with century (1987)
%Z  |  Time zone abbreviation as a character string (empty if not available)

## Dates with the Lubridate Package

The *lubridate* package was developed to make date conversions faster and simpler.  This package contains a few basic commands that will convert all of the most common date formats without the user having to specify their unique data format.  

The basic *lubridate* date conversions are ``ymd`` (year-month-day), ``mdy`` (month-day-year), and ``dmy`` (day-month-year). 

We've illustrated how to use these functions below:

```{r,warning=FALSE,message=FALSE}
library(lubridate)
ymd("2016-02-07", "2016-04-02","2016-06-28")


```


Now we'll use ``mdy`` to convert in a different format. 


```{r}

mdy("02/07/2016", "04/02/2016","06/28/2016")

```

To show the flexibility of this code, we'll do a final example with ``dmy`` used on a different format data:

```{r}
dmy("1jan16", "1nov15","15mar17")

```

The *lubridate* commands can be expanded to include hour-minute-seconds as well

```{r}

ymd_hms("2016-10-10 17:46:52", "2016-11-14 12:04:05", "2016-10-22 22:44:58")
```

If you have times in different time zones, you can add a time zone parameter:

```{r}
ymd_hms("2016-10-10 17:46:52", tz="Pacific/Aukland")
```

Note: UTC and GMT are Greenwich Mean Time (also known as "Zulu" time).


##POSIXct and POSIXlt

To fully understand how dates work in R, you will need study and understand the POSIXct and POSIXlt classes.  You can learn more about these by typing ``?POSIXct`` or ``?POSIXlt`` respectively.

